Realizability-Preserving DG-IMEX Method for the Two-Moment Model of Fermion Transport \tnoteref{support}

### 2018-08-07 Documents/M1/realizableFermionicM1/sections/intro.tex
Work on intro

In this paper we design numerical methods to solve the two-moment model for transport of particles governed by Fermi-Dirac statistics.  
The numerical method is based on the Discontinuous Galerkin (DG) method for spatial discretization, Implicit-Explicit (IMEX) methods for time integration, and preserves physical constraints of the underlying model.  
The latter property is achieved by considering the spatial and temporal discretization together with the closure procedure for the two-moment model.

In many applications, the particle mean free path exceeds other characteristic length scales in the system under consideration (i.e., the Knudsen number exceeds unity), and non-equilibrium effects become important.  
In these situations, a kinetic description may be required, and the particle distribution function $f(\vect{p},\vect{x},t)$ is solved for.  
The distribution function, a phase space density depending on momentum $\vect{p}$ and position $\vect{x}$, gives at time $t$ the number of particles in the phase space volume element $d\vect{p}\,d\vect{x}$; i.e., $dN=f\,d\vect{p}\,d\vect{x}$.  
The evolution of the distribution function is governed by the Boltzmann equation, which states a balance between phase space advection and collisions.

Solving the Boltzmann equation for $f$ numerically is challenging, in part due to the high dimensionality of phase space.  
To reduce the dimensionality of the problem and make it more computationally tractable, one may instead solve for a finite number of angular moments $\{m^{(k)}\}_{k=0}^{N}$ of the distribution function, defined as
\begin{equation}
  m^{(k)}(\varepsilon,\vect{x},t)=\f{1}{4\pi}\int_{\bbS^{2}}f(\omega,\varepsilon,\vect{x},t)\,g^{(k)}(\omega)\,d\omega,
\end{equation}
where $\varepsilon=|\vect{p}|$, $\omega$ is a point on the unit sphere $\bbS^{2}$ indicating the particle propagation direction, and $g^{(k)}$ are momentum space angular weighing functions.  
In problems where collisions are sufficiently frequent, solving this \emph{truncated moment problem} can give significant reductions in the computational cost since only a few moments are needed to represent the solution accurately.  
In problems where collisions do not sufficiently isotropize the angular dependence of the distribution function, more moments are needed to accurately represent the solution.  
In the so-called two-moment model, angular moments representing the particle density and flux vector are solved for.

When solving the truncated moment problem, the equation governing the evolution of the $N$-th order moment $m^{(N)}$ contains higher order moments $\{m^{(k)}\}_{k=N+1}^{M}$ ($M>N$), which must be specified to form a closed system of equations.  
For the two-moment model, the symmetric rank-two Eddington tensor (proportional to the pressure tensor) must be specified.  
Solutions to the closure problem include setting $m^{(k)}=0$, for $k>N$, Eddington approximation, Kershaw-type closure, and maximum entropy closure.  
The closure procedure often results in a system of nonlinear hyperbolic conservation laws, which can be solved using suitable numerical methods.  
One challenge in solving the closure problem is constructing a sequence of moments that are consistent with a positive distribution function, which typically implies algebraic constraints on the moments.  
Moments satisfying these constraints are called \emph{realizable moments}.

When solving the Boltzmann equations approximately by a truncated moment model, maintaining realizable moments is necessary in order to ensure the well-posedness of the closure procedure, and may also be necessary to guarantee other desirable properties of the model, including hyperbolicity and entropy dissipation.  
Failure to maintain moment realizability in a numerical model may then require ad hoc post-processing steps with undesirable consequences such as loss of conservation.  
For the two-moment model for particles governed by Maxwell-Boltzmann statistics (``classical'' particles with $f\ge0$), the particle density is nonnegative, the magnitude of the flux vector is bounded by the particle density, and there are further constraints on the components of the Eddington tensor \cite{levermore_1984}.  
Furthermore, the realizable set of moments spanned by the particle density and flux vector constitute a convex cone \cite{olbrant_etal_2012}.

>>>>

In this paper we design numerical methods to solve the two-moment model governing transport of particles obeying by Fermi-Dirac statistics (e.g., neutrinos), with the ultimate target being nuclear astrophysics applications (e.g., neutrino transport in core-collapse supernovae and compact binary mergers).  
The numerical method is based on the discontinuous Galerkin (DG) method for spatial discretization, implicit-explicit (IMEX) methods for time integration, and is designed to preserve physical constraints of the underlying model.  
The latter property is achieved by considering the spatial and temporal discretization together with the closure procedure for the two-moment model.

In many applications, the particle mean free path is comparable to or exceeds other characteristic length scales in the system under consideration (i.e., the Knudsen number exceeds unity), and non-equilibrium effects may become important.  
In these situations, a kinetic description may be required, where particle distribution function $f(\vect{p},\vect{x},t)$ is solved for.  
The distribution function, a phase space density depending on momentum $\vect{p}$ and position $\vect{x}$, gives at time $t$ the number of particles in the phase space volume element $d\vect{p}\,d\vect{x}$ (i.e., $dN=f\,d\vect{p}\,d\vect{x}$).  
The evolution of the distribution function is governed by the Boltzmann equation, which states a balance between phase space advection and collisions (see, e.g., \cite{braginskii_1965,chapmanCowling_1970,lifshitzPitaevskii_1981}).

Solving the Boltzmann equation numerically for $f$ is challenging, in part due to the high dimensionality of phase space.  
To reduce the dimensionality of the problem and make it more computationally tractable, one may instead solve for a finite number of angular moments $\{m^{(k)}\}_{k=0}^{N}$ of the distribution function, defined as
\begin{equation}
  m^{(k)}(\varepsilon,\vect{x},t)=\f{1}{4\pi}\int_{\bbS^{2}}f(\omega,\varepsilon,\vect{x},t)\,g^{(k)}(\omega)\,d\omega,
\end{equation}
where $\varepsilon=|\vect{p}|$ is the particle energy, $\omega$ is a point on the unit sphere $\bbS^{2}$ indicating the particle propagation direction, and $g^{(k)}$ are momentum space angular weighing functions.  
In problems where collisions are sufficiently frequent, solving a \emph{truncated moment problem} can provide significant reductions in computational cost since only a few moments are needed to represent the solution accurately.  
On the other hand, in problems where collisions do not sufficiently isotropize the angular dependence of the distribution function, more moments may be needed.  
In the two-moment model considered here, angular moments representing the particle density and flux (or energy density and momentum) are solved for.  
Two-moment models for relativistic systems appropriate for nuclear astrophysics applications have been discussed in, e.g., \cite{lindquist_1966,andersonSpiegel_1972,thorne_1981,shibata_etal_2011,cardall_etal_2013a}.  
However, in this paper, for simplicity (and clarity), we consider a non-relativistic model, leaving extensions to relativistic systems for future work.

When solving a truncated moment problem, the equation governing the evolution of the $N$-th order moment $m^{(N)}$ contains higher order moments $\{m^{(k)}\}_{k=N+1}^{M}$ ($M>N$), which must be specified to form a closed system of equations.  
For the two-moment model, the symmetric rank-two Eddington tensor (proportional to the pressure tensor) must be specified.  
Solutions to the closure problem include setting $m^{(k)}=0$, for $k>N$, Eddington approximation, Kershaw-type closure, and maximum entropy closure.  
The closure procedure often results in a system of nonlinear hyperbolic conservation laws, which can be solved using suitable numerical methods.  
One challenge in solving the closure problem is constructing a sequence of moments that are consistent with a positive distribution function, which typically implies algebraic constraints on the moments.  
Moments satisfying these constraints are called \emph{realizable moments}.

When solving the Boltzmann equations approximately by a truncated moment model, maintaining realizable moments is necessary in order to ensure the well-posedness of the closure procedure, and may also be necessary to guarantee other desirable properties of the model, including hyperbolicity and (mathematical) entropy dissipation.  
Failure to maintain moment realizability in a numerical model may then require ad hoc post-processing steps with undesirable consequences such as loss of conservation.  
For the two-moment model for particles governed by Maxwell-Boltzmann statistics (``classical'' particles with $f\ge0$), the particle density is nonnegative, the magnitude of the flux vector is bounded by the particle density, and there are further constraints on the components of the Eddington tensor \cite{levermore_1984}.  
Furthermore, the realizable set of moments spanned by the particle density and flux vector constitute a convex cone \cite{olbrant_etal_2012}.
